---
title: "Human Activity Recognition from Wearable Sensors"
output: html_document
---

```{r setup, include=FALSE}
library(caret)
library(ranger)
library(ggplot2)
library(dplyr)
```

## Data

This project uses the Weight Lifting Exercise Dataset from the Practical Machine Learning course. Files are stored under `data/` and downloaded automatically if they are missing.

```{r}
set.seed(1821)

data_dir <- "data"
dir.create(data_dir, showWarnings = FALSE)

training_path <- file.path(data_dir, "pml-training.csv")
testing_path <- file.path(data_dir, "pml-testing.csv")

training_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testing_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

if (!file.exists(training_path)) {
  download.file(training_url, training_path, mode = "wb")
}
if (!file.exists(testing_path)) {
  download.file(testing_url, testing_path, mode = "wb")
}

na_values <- c("NA", "#DIV/0!", "")
raw_training <- read.csv(training_path, na.strings = na_values, check.names = FALSE)
raw_testing <- read.csv(testing_path, na.strings = na_values, check.names = FALSE)

stopifnot("classe" %in% names(raw_training))
```

## Training-derived feature selection

Feature-selection decisions are learned from the labelled training dataset and then applied unchanged to the quiz/test dataset. This avoids independently filtering the two files into different schemas.

```{r}
metadata_columns <- c(
  "X", "user_name", "raw_timestamp_part_1", "raw_timestamp_part_2",
  "cvtd_timestamp", "new_window", "num_window", "problem_id"
)

candidate_predictors <- setdiff(names(raw_training), c("classe", metadata_columns))

# Retain predictors that are not overwhelmingly missing in training data.
missing_fraction <- vapply(
  raw_training[candidate_predictors],
  function(x) mean(is.na(x)),
  numeric(1)
)
selected_predictors <- candidate_predictors[missing_fraction <= 0.95]

# Remove near-zero-variance predictors using training data only.
nzv <- nearZeroVar(raw_training[selected_predictors])
if (length(nzv) > 0) {
  selected_predictors <- selected_predictors[-nzv]
}

missing_in_quiz <- setdiff(selected_predictors, names(raw_testing))
if (length(missing_in_quiz) > 0) {
  stop(
    "Quiz data are missing predictors selected from training data: ",
    paste(missing_in_quiz, collapse = ", ")
  )
}

model_data <- raw_training %>%
  select(all_of(selected_predictors), classe) %>%
  mutate(classe = factor(classe))

quiz_data <- raw_testing %>%
  select(all_of(selected_predictors))

cat("Predictors retained:", length(selected_predictors), "\n")
```

## Hold-out evaluation

A stratified hold-out split is created before model fitting. Cross-validation is performed only within the training partition.

```{r}
set.seed(1821)
train_index <- createDataPartition(model_data$classe, p = 0.75, list = FALSE)
train_data <- model_data[train_index, ]
test_data <- model_data[-train_index, ]

control <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  savePredictions = "final"
)

set.seed(1821)
rf_model <- train(
  classe ~ .,
  data = train_data,
  method = "ranger",
  trControl = control,
  tuneLength = 5,
  metric = "Accuracy",
  num.trees = 500,
  importance = "permutation"
)

rf_model
```

## Held-out performance

```{r}
holdout_predictions <- predict(rf_model, newdata = test_data)
holdout_confusion <- confusionMatrix(holdout_predictions, test_data$classe)
holdout_confusion
```

## Variable importance

```{r}
importance <- varImp(rf_model)
plot(importance, top = 20, main = "Top predictor importance")
```

## Final model and quiz predictions

After the hold-out result has been inspected, a final model is fitted on all labelled observations using the same predictor schema. This model is used only for the course quiz/test rows.

```{r}
set.seed(1821)
final_model <- train(
  classe ~ .,
  data = model_data,
  method = "ranger",
  trControl = control,
  tuneLength = 5,
  metric = "Accuracy",
  num.trees = 500,
  importance = "permutation"
)

quiz_predictions <- predict(final_model, newdata = quiz_data)
quiz_predictions
```

## Notes

- The old exploratory course submission is retained under `archive/` for provenance.
- The root HTML report is a historical render and may not reflect this audited source.
- Normality testing is intentionally omitted because it is not an assumption of the Random Forest model used here.

## Session information

```{r}
sessionInfo()
```
